Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales

### 2019-10-22 reports/article.Rmd
last version before asking for external validation

---
title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational register data from Sweden, England and Wales"
author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Erik Lenguerrand^5^, Ashley Blom^6^, Mark Wilkinson^7^, Adrian Sayers^8^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
date: '`r Sys.Date()`'
bibliography: P:/library.bib
csl: the-lancet.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
---

@@

We also built an online web calculator available at https://erikbulow.shinyapps.io/thamortpred/. Patients can use it to easily estimate their own risk of mortality. This estimate can be used in a discussion of shared decision making on whether to operate or not.

@@

The resulting "BRL reduced" model can predict the probability of death within 90 days as: $p = 1 / [1 +\exp(\beta X)]$ where `r coefs_form`
This formula is valid for patients within the observed age range (`r paste(range(df$P_Age), collapse = " - ")` years) with observed combinations of diseases.

@@

For example a `r min(df$P_Age)` year old woman with ASA grade = 1 and none of the important comorbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery. Another woman, `r quantile(df$P_Age, .25)` years old (the first age quantile), have an elevated risk of `r get_risk("healthy_female")`. A `r quantile(df$P_Age, .75)` years old man (the third age quantile) with ASA grade = 3 and a previous heart condition would have a risk of `r get_risk("sicker_man")`. If that man was instead 99 years old (the maximum observed age), with cancer as well, his risk would be `r get_risk("sickest_man")`. Note however that observed covariate patters with predicted probabilities above 3% were rare. Only `r mean(pred_probs > 3) * 100`% (`r sum(pred_probs > 3)` patients) had those (Figure \@ref(fig:sep)). Estimated mortality risks above 3% are therefore subject to extrapolation. Some risk calculators ignore this problem,[@Bozic2013a] but we think this should be acknowledged.

Obesity in the "BRL reduced" model was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but is relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We note that the proportion of patients with BMI above 30, the definition of obisity (`r mean(df$P_BMI > 30, na.rm = TRUE) * 100`%) is much higher than the proportion of patients with ICD-10 = E66* (obesity) from the national patient register (`r mean(df$c_obesity) * 100`%). This proportion corresponds to patients with BMI of at least `r which.min(map_dbl(1:50, ~mean(df$P_BMI > ., na.rm = TRUE)) > mean(df$c_obesity))`. Thus, obesity might be indicated in the risk calculator only if the condition is severe.

@@

It is known that male sex is associated with earlier deaths and that the remaining life span will decrease with increased age. It is less obvious that this relation must be linear. We used restricted cubic splines to allow a more flexible relation, but found that a linear relationship was equally good. Our model includes ASA grade, which is routinely assessed pre-operatively in most developed countries. It is however known to have a high degree of internal variability.[@Haynes1995] It has previously been compared to the Charlson comorbidity index, but not with respect to mortality after THA.[@Whitmore2014; @Kork2015] Patients with ASA grade 4-6 were excluded since those categories describe severe disease, moribund and brain-dead individuals. Patients with those conditions are unlikely to get elective THA.

>>>>

---
title: "Prediction of 90-day mortality after Total Hip Arthroplasty: a simplified and externally validated model based on observational registry data from Sweden, England and Wales"
author: Anne Garland^1,2,3^, Erik Bulow^2,4^, Erik Lenguerrand^5^, Ashley Blom^6^, Mark Wilkinson^7^, Adrian Sayers^8^, Ola Rolfson^2,4^, Nils P. Hailer^1,2^
date: '`r Sys.Date()`'
bibliography: P:/library.bib
csl: the-lancet.csl
output: 
  bookdown::word_document2:
    reference_docx: article.dotx
    toc: false
    df_print: kable
---

@@

We also built an online web calculator available at https://erikbulow.shinyapps.io/thamortpred/.

@@

The resulting "BRL reduced" model can predict the probability of death within 90 days as: $p = 1 / [1 +\exp(-\beta X)]$ where `r coefs_form`
This formula is valid for patients within the observed age range (`r paste(range(df$P_Age), collapse = " - ")` years) with observed combinations of diseases.

@@

For example a `r min(df$P_Age)` year old woman with ASA grade = 1 and none of the important comorbidities would have a `r get_risk("baseline")` risk to die within 90 days of surgery. Another woman, `r quantile(df$P_Age, .25)` years old (the first age quantile), have an elevated risk of `r get_risk("healthy_female")`. A `r max(df$P_Age)` years old man (the maximum observed age) with ASA grade = 3 and cancer, would have a risk of `r get_risk("sickest_man")`. Note however that observed covariate patters with predicted probabilities above 3% were rare. Only `r mean(pred_probs > 3) * 100`% (`r sum(pred_probs > 3)` patients) had those (Figure \@ref(fig:sep)). Estimated mortality risks above 3% are therefore subject to extrapolation. Some risk calculators ignore this problem,[@Bozic2013a] but we think this should be acknowledged.

Obesity in the "BRL reduced" model was not statistically significant by itself ($p =$ `r subset(brlasso_tbl_coefs, term == "Obesity", "p", drop = TRUE)`) but is relevant as a predictor due to unobserved heterogeneity.[@Mood2010] We note that the proportion of patients with BMI above 30, the definition of obesity (`r mean(df$P_BMI > 30, na.rm = TRUE) * 100`%) is much higher than the proportion of patients with ICD-10 = E66* (obesity) from the National Patient Register (`r mean(df$c_obesity) * 100`%). This proportion corresponds to patients with BMI of at least `r which.min(map_dbl(1:50, ~mean(df$P_BMI > ., na.rm = TRUE)) > mean(df$c_obesity))`. Thus, obesity might be indicated in the risk calculator only if the condition is severe.

@@

It is known that male sex is associated with earlier deaths and that the remaining life span will decrease with increased age. It is less obvious that this relation must be linear. We used restricted cubic splines to allow a more flexible relation, but found that a linear relationship was equally good. Our model includes ASA grade, which is routinely assessed pre-operatively in most developed countries. It is however known to have a high degree of internal variability.[@Haynes1995] It has previously been compared to the Charlson comorbidity index, but not with respect to mortality after THA.[@Whitmore2014; @Kork2015] Patients with ASA grade 4-6 were excluded since those categories describe severe disease, moribund and brain-dead individuals. We suspect that patients with those ASA grades might be misclassified. Their true grade could either be lower, or they could have recieved THA due to a femoral hip fracture instead of osteoarthritis.

### 2019-10-22 reports/figs_tabs.Rmd
last version before asking for external validation

```{r sep, fig.cap = "Most patients survived more than 90 days after THA, wherefore the blue bars totally dominate the histogram (upper panel). A standardized density plot however reveals that patients who died within 90 days (pink) were, on average, estimated to have a higher probability to do so (lower panel). There were very few observations with covariate patterns resultnig in death probabilities higher than 3%."}
knitr::include_graphics("../graphs/brlasso_separation_auc.png")
```

@@

```{r calibration, fig.cap = "Calibration belts at 80% and 95% (light and dark grey). The predicted probabilities do not deviate from the observed ones by any statistical significance. The risk of over-estimation is however clear above 3%."}
knitr::include_graphics("../graphs/calibration_belt.png")
```

>>>>

```{r sep, fig.cap = "Most patients survived more than 90 days after THA, wherefore the blue bars totally dominate the histogram (upper panel; note the square-root scale on the y-axis, that was necessary to distinguish the pink bars from the horisontal line). A normalized density plot however reveals that patients who died within 90 days (pink) were, on average, estimated to have a higher probability to do so (lower panel). There were very few observations with covariate patterns resultnig in death probabilities higher than 3%."}
knitr::include_graphics("../graphs/brlasso_separation_auc.png")
```

@@

```{r calibration, fig.cap = "Calibration belts based on deciles of fitted probabilities and loess smoothers at 80% and 95% confidence levels (light and dark grey). The predicted probabilities do not deviate much from the observed ones up to approximately 3%. A predicted risk of 6% however could correspond to observed probabilities between 2 and 8% (the width of the 95% belt). The higher the predicted probability gets, the more it over-estimates the observed probabilities."}
knitr::include_graphics("../graphs/calibration_belt.png")
```
